---
title: "Data Cleaning"
author: "Kylie Newcomer"
format: html
execute: 
  message: false
  warning: false
---
# Load necessary packages and data
```{r}
library(tidyverse)
library(here)
library(janitor)
```

```{r}
hal <- read_csv(here("data","hal_trawl_fish.csv")) %>% 
  clean_names()

trawl_info <- read_csv(here("data", "hal_trawl_info.csv"))%>% 
  clean_names()

other_sp <- read_csv(here("data", "hal_trawl_sp.csv")) %>% 
  clean_names()

sp_length <- read_csv(here("data", "hal_trawl_sp_length.csv")) %>% 
  clean_names()

location <- read_csv(here("data", "hal_trawl_loc_meta.csv")) %>% 
  clean_names()

```
# Clean trawl information
I started by removing the end columns with no values. I also wanted to convert the date column to a date time data type. There were also two columns that should be numeric but weren't due to inconsistent data entry. I forced those values to NA's to create numeric columns. There were also filler rows and data from another study (Maturity) in the dataset that the data owner requested that I remove.
```{r}
trawl <- trawl_info %>% 
  select(-c(location, x27:x31)) %>% 
  mutate(date = as.Date(date, format = "%m/%d/%y"),
         halibut_100m2 = as.numeric(halibut_100m2),
         start_depth = as.numeric(start_depth)) %>% 
  filter(!is.na(date) & loc != "M" & total_trawl_count != "Maturity")
```

In order to ensure each trawl ID is unique, I will filter all other data based on `trawl_id` from the cleaned trawl data.
```{r}
true_trawl_id <- trawl %>% 
  select(trawl_id)
```

# Clean other species observations
This data has many observed species, and with that, there was lots of inconsistent naming, spelling errors, and lots of questions marks for unidentified species. I had to manually address all spelling or inconsistent names. I replaced species with question marks to a larger taxonomic classification for both the common name and scientific name. There was also lots of incorrect use of "sp." vs "spp.", so that was corrected at the end.
```{r}
species_fixed <- other_sp  %>% 
  filter(trawl_id %in% true_trawl_id$trawl_id) %>% 
  # Some of the time target shrimp was recorded in the common_name column - other times it was in the 'other_species' column
  # Add Target shrimp to common name and remove from other_species
  mutate(common_name = case_when(other_species == "Target Shrimp" ~ "Target shrimp",
                                 TRUE ~ common_name),
         other_species = case_when(other_species == "Target Shrimp" ~ NA,
                                   TRUE ~ other_species),
         other_species = str_replace(other_species, "\\?", ""),
         # Find all fish that were not identified to species
         # Assign broader taxonomic classification to each
         # Correct spelling inconsistencies
         common_name = str_to_sentence(replace_values(common_name,
                                 c("White croaker?", "White Seabass?", "Unidentified Croaker", "Croacker sp.") ~ "Croaker sp.",
                                 "Rockfish?" ~ "Rockfish sp.",
                                 "Topsmelt?" ~ "Silverside sp.",
                                 "Barred/Spotted Sand Bass?" ~ "Bass sp.",
                                 "Sea Squirts?" ~ "Sea Squirt",
                                 c("Hornyhead Turbot?", "Hornyhead turbot/ Curlfin?", "Turbot/sole Sp.") ~ "Turbot/Sole sp.",
                                 "Horneyhead Turbot" ~ "Hornyhead turbot",
                                 "Vermillion Rockfish" ~ "Vermilion rockfish",
                                 "Unidentified Perch" ~ "Surfperch sp.",
                                 "Snail" ~ "Snails",
                                 "C-O Sole" ~ "C-O Turbot",
                                 "Basket weave cusk eel" ~ "Basketweave cusk eel",
                                 "Hornshark" ~ "Horn shark",
                                 "Lizardfish" ~ "California lizardfish",
                                 "Round Ray" ~ "Round stingray",
                                 "Larval Queenfish" ~ "Queenfish",
                                 "Sand dabs" ~ "Sand dab sp.",
                                 "Pipefish" ~ "Pipefish sp.",
                                 # not defined species
                                 "White Anchovy" ~ "Anchovy sp.",
                                 "Translucent flatfish" ~ "Flatfish sp.",
                                 "Urchin" ~ "Sea urchin")),
         # Add or or correct scientific name for newly classified species
         other_species = case_when(common_name == "Croaker sp." ~ "Sciaenidae sp.",
                                   common_name == "Rockfish sp." ~ "Sebastes sp.",
                                   common_name == "Silverside sp." ~ "Atherinopsidae sp.",
                                   common_name == "Bass sp." ~ "Paralabrax sp.",
                                   common_name == "Sea Squirt" ~ "Invertebrate",
                                   common_name == "Turbot/sole sp." ~ "Pleuronectidae sp.",
                                   common_name == "Surfperch sp." ~ "Embiotocidae sp.",
                                   common_name == "Sand dab sp." ~ "Citharichthys sp.",
                                   # Correct scientific names for inconsistently named species
                                   common_name == "Deep-body anchovy" ~ "Anchoa compressa",
                                   common_name == "Navanax" ~ "Invertebrate",
                                   common_name == "Black sea hare" ~ "Invertebrate",
                                   common_name == "Pacific sand dab" ~ "Citharichthys sordidus",
                                   common_name == "Topsmelt" ~ "Atherinops affinis",
                                   common_name == "Clingfish" ~ "Gobiesocidae sp.",
                                   common_name == "Diamond turbot" ~ "Hypsopsetta guttulata",
                                   common_name == "Salema" ~ "Haemulon californiensis",
                                   common_name == "Walleye surfperch" ~ "Hyperprosopon argenteum",
                                   common_name == "Pipefish sp." ~ "Syngnathus sp.",
                                   common_name == "Anchovy sp." ~ "Engraulis sp.",
                                   common_name == "Flatfish sp." ~ "Flatfish sp.",
                                   # Identify invert observations as Invertebrate classification
                                   is.na(other_species) ~ "Invertebrate",
                                   TRUE ~ other_species),
         # Move presence/absence note for invertebrates to `notes` column rather than the `count`
         notes = case_when(other_species == "Invertebrate" ~ "Present",
                           TRUE ~ notes),
         common_name = case_when(common_name == "Octopus" ~ common_name,
                                 other_species == "Invertebrate" ~ str_replace(common_name, "s$", ""),
                                 TRUE ~ common_name),
         # Remove or replace any non-numeric values in `count` column
         count = as.numeric(case_when(
           other_species == "Invertebrate" ~ NA,
           count == "~500" ~ "500",
           count == "~200" ~ "200",
           TRUE ~ count))
         ) %>% 
  # Update `other_species` column to `scientific_name`
  rename(scientific_name = other_species) %>% 
  # Remove spacer rows and unknown species (confirmed with data owner that this should be removed)
  filter(!is.na(common_name) & !str_detect(common_name, "pusgill"))

# Change inconsistency with spp. vs sp.
species_fixed$common_name <- str_squish(str_replace(species_fixed$common_name, "spp.", "sp."))
species_fixed$scientific_name <- str_replace(species_fixed$scientific_name, "spp.", "sp.")

```

# Correct species in lengths
Same as the species counts, there were naming inconsistencies that I had to correct.
```{r}
other_sp_length <- sp_length %>% 
  # Filter to remove blank rows and maturity trawls
  filter(!is.na(species) & trawl_id %in% true_trawl_id$trawl_id) %>% 
  mutate(species = str_to_sentence( # Correct species names
           replace_values(species, c("White croaker?", "White Seabass?", "White Sea Bass?", "Unidentified Croaker") ~ "Croaker sp.",
                          "White Sea Bass" ~ "White seabass",
                          c("Barred Surfperch?", "Perch", "Unidentified perch") ~ "Surfperch sp.",
                          "Ca Scorpionfish" ~ "California scorpionfish",
                          "Horney Turbot" ~ "Hornyhead turbot",
                          "Jack Smelt" ~ "Jacksmelt",
                          "Top Smelt" ~ "Topsmelt",
                          "Northern Anchovy?" ~ "Anchovy sp.",
                          "Pacific pampano" ~ "Pacific pompano",
                          "Rockfish?" ~ "Rockfish sp.",
                          "Dwarf Surfperch" ~ "Dwarf Perch",
                          "Zebra Surfperch?" ~ "Zebraperch"
    
  ))
)
```

# Clean Halibut data
This data did not have too many inconsistencies to address. The main thing was filling in the length bins for the most recent halibut entries. I first converted each length to inches then matched to the correct bin. There was also inconsistent noting of if the old or new net was used, which began in 2024. 
```{r}
halibut_bins <- hal %>% 
  # Convert length to inches
  mutate(length_in = halibut_total_length_mm * 0.03937,
         # Assign bins
         length_bin = case_when(
           length_in > 21 ~ ">21",
           length_in >= 18 ~ "18-21",
           length_in >= 15 ~ "15-17",
           length_in >= 12 ~ "12-14",
           length_in >= 9 ~ "9-11",
           length_in >= 7 ~ "7-8",
           length_in >= 5 ~ "5-6",
           length_in >= 3 ~ "3-4",
           length_in >= 1 ~ "1-2",
           TRUE ~ "<1"
         ),
         # Remove net type from condidition notes
         condition_notes = replace_values(condition_notes,
                                          "Old Net" ~ NA,
                                          "New Net" ~ NA),
         # Standardize positive otolith extraction distiction
         otoliths = replace_values(otoliths,
                                   "Yes" ~ "Y")) %>% 
  # Remove length in inches
  select(-length_in) %>% 
  # Remove maturity trawl halibut and blank rows
  filter(trawl_id %in% true_trawl_id$trawl_id & trawl_id != "#N/A")

```

# Clean Site code data
For the site codes, the only issue was the "Maturity" sites, which were trawl sites only visited to collect maturity data. Those sites were removed.
```{r}
site <- location %>% 
  filter(location_abbreviation != 'Maturity')
```

# Write clean data to CSVs
```{r}
write_csv(other_sp_length, here::here("data", "processed", "other_sp_length.csv"))
```

```{r}
write_csv(other_species_names, here::here("data", "processed", "species_key.csv"))
```

```{r}
write_csv(halibut_bins, here::here("data", "processed", "halibut_clean.csv"))
```

```{r}
write_csv(species_fixed, here::here("data", "processed", "species_fixed.csv"))
```

```{r}
write_csv(trawl, here::here("data", "processed", "trawl.csv"))
```

```{r}
write_csv(site, here::here("data", "processed", "site.csv"))
```

